---
title: "SEIR Modeling"
author: "Rob Carnell"
date: "April 6, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(deSolve)
require(ggplot2)
require(reshape2)
require(magrittr)
```

## Solving Differential Equations

### Separable, Homogenous System of ODEs

- **Separable**: each differential equation is on an expression of one function of time
- **Homogeneous**: only the functions and their derivatives are present.  No other functions of t
- **Ordinary**: Only a function of t and derivates of t

## Example ODE

Imagine a process where water flows from bucket A to B to C at different rates.
Plot the volume in each bucket with time.

- $V_A$, $V_B$, $V_C$ are the volumes of water in the buckets ($L$).  $V_A \ge 0$, $V_B \ge 0$, $V_C \ge 0$
- $k > 0$, $h > 0$, $k > h$ are the rates of water flow ($L/s$)
- $V_A(0) = 5 L$, $V_B(0) = V_C(0) = 0$

```{r water_diagram, echo=FALSE, fig.height=3, fig.width=5}
temp <- DiagrammeR::grViz(diagram = "
  digraph boxes_and {
    graph[overlap = true, fontsize = 10, rankdir = LR]

    node [shape = box] A; B; C

    A->B [label='k']
    B->C [label='h']
  }")
temp
rm(temp) # helps to prevent "out of memory errors"

```

## Equations

$$\frac{dV_A}{dt} = -k$$

$$\frac{dV_B}{dt} = k - h$$

$$\frac{dV_C}{dt} = h$$

## Equations (2)

Integrating each expression yields:

$$\int_0^t \frac{dV_A}{dt}dt = V_A(t)-V_A(0) = -\int_0^t k dt = -kt + k(0) = -kt$$

or

$$\int \frac{dV_A}{dt}dt = V_A(t) = -\int k dt = -kt + C$$

## Equations (3)

$$V_A(0) = -k(0) + C,\ \ \ C = V_A(0)$$

$$V_A(T_A) = 0 = -kT_A + V_A(0), \ \ \ T_A = \frac{V_A(0)}{k}$$


$$V_A(t) = -kt + V_A(0), \ \ \ 0 \le t \le T_A$$

## Equations (4)

$$V_B(t) = (k - h)t + V_B(0), \ \ \ 0 \le t \le T_A$$

$$V_B(t) = -ht + (k-h)T_A + V_B(0), \ \ \ T_A \lt t \le T_B$$

$$T_B = \frac{kT_A + V_B(0)}{h}$$

$$V_C(t) = ht + V_C(0), \ \ \ 0 \le t \le T_C$$

$$T_C = \frac{V_A(0) - V_C(0)}{h}$$

## Bucket Code

```{r simple, echo=TRUE}
VA <- function(t, k, VA0) ifelse(t > VA0 / k, 0, -k*t + VA0)
VB <- function(t, k, h, VA0, VB0)
{
  TA <- VA0 / k
  TB <- (k * TA + VB0) / h
  ifelse(t < TB, 
         ifelse(t < TA, (k - h) * t + VB0, -h * t + k*TA + VB0), 
         0)
}
VC <- function(t, h, VA0, VC0) 
{
  TC <- (VA0 - VC0) / h 
  ifelse(t < TC, h*t + VC0, h*TC + VC0)
}
```

## Bucket Figure

```{r simple_figure, echo = FALSE}
plot_data <- data.frame(t = seq(0, 6, by = 0.1))
plot_data$VA <- VA(plot_data$t, 3, 5)
plot_data$VB <- VB(plot_data$t, 3, 1, 5, 0)
plot_data$VC <- VC(plot_data$t, 1, 5, 0)
plot_data <- plot_data %>% reshape2::melt(id = "t")

g <- ggplot(plot_data, aes(x = t, y = value, col = variable)) +
  geom_line() +
  labs(x = "time", y = "Volume", col = "Bucket")
plot(g)
```

## Linear Homogenous System of ODEs

Brine Tank Cascade.  Water enters tank A and moves between tanks at a constant rate
so that the tank volumes are constant.  Each tank has a different amount of salt
in solution.

- $V_A$, $V_B$, $V_C$ are the volumes of water in the buckets (L).  $V_A \ge 0$, $V_B \ge 0$, $V_C \ge 0$
- $r > 0$ is the rates of water flow (L/s)
- $x_A(t)$ is the amount of salt in bucket A (kg)

```{r salt_diagram, echo=FALSE, fig.height=3, fig.width=5}
temp <- DiagrammeR::grViz(diagram = "
  digraph boxes_and {
    graph[overlap = true, fontsize = 10, rankdir = LR]

    node [shape = box] Fresh_Water; A; B; C; Drain

    Fresh_Water->A [label='r']
    A->B [label='r']
    B->C [label='r']
    C->Drain [label='r']
  }")
temp
rm(temp) # helps to prevent "out of memory errors"
```

## Equations

$$x'_A = \frac{-r}{V_A}x_A = -a x_A$$

$$x'_B = \frac{r}{V_A}x_A - \frac{r}{V_B}x_B = a x_A - b x_b$$

$$x'_C = \frac{r}{V_B}x_B - \frac{r}{V_C}x_C = b x_B - c x_C$$

## Laplace transforms

$$L[x_A(t)] = A(s)$$

$$L[x'_A(t)] = sA(s) - x_A(0)$$

Applied to our differential equations...

$$sA(s) - x_A(0) = -aA(s)$$
$$sB(s) - x_B(0) = aA(s) - bB(s)$$
$$sC(s) - x_C(0) = bB(s) - cC(s)$$

## Some algebra

$$A(s) = \frac{x_A(0)}{s+a}$$

$$x_A(t) = x_A(0)e^{-at}$$

$$B(s) = \frac{ax_A(0)}{(s+a)(s+b)} + \frac{x_B(0)}{s+b}$$

$$= \frac{-a x_A(0) / (a-b)}{s+a} + \frac{a x_A(0) / (a-b)}{s+b} + \frac{x_B(0)}{s+b}$$
$$x_b(t) = \frac{-ax_A(0)}{a-b}e^{-at} + \frac{ax_A(0)}{a-b}e^{-bt} + x_B(0)e^{-bt}$$

## Some Harder Algebra {.smaller}

$$C(s) = \frac{-ab x_A(0) / (a-b)}{(s+a)(s+c)} + \frac{ab x_A(0) / (a-b)}{(s+b)(s+c)} + \frac{bx_B(0)}{(s+b)(s+c)} + \frac{x_C(0)}{s+c}$$

$$C(s) = \frac{-w}{(s+a)(s+c)} + \frac{w}{(s+b)(s+c)} + \frac{g}{(s+b)(s+c)} + \frac{x_C(0)}{s+c}$$

$$C(s) = \frac{w/(a-c)}{s+a} + \frac{-w/(a-c)}{s+c} + \frac{-w/(b-c)}{s+b} + \frac{w/(b-c)}{s+c} + $$

$$\frac{-g/(b-c)}{s+b} + \frac{g/(b-c)}{s+c} + \frac{x_c(0)}{s+c}$$

$$x_C(t) = \frac{abx_A(0)}{(a-b)(a-c)}e^{-at} + \frac{-abx_A(0)/(a-b) - bx_B(0)}{(b-c)}e^{-bt} +$$

$$\frac{Stuff}{Other\ Stuff}e^{-ct}$$

## Figure

```{r salt, echo=FALSE}
VA <- 20
VB <- 40
VC <- 60
r <- 10
a <- r / VA
b <- r / VB
cc <- r / VC
xA0 <- 5
xB0 <- 7
xC0 <- 11
w <- a*b*xA0/(a - b)
g <- b*xB0
pd <- data.frame(t = seq(0, 50, by = 0.1))
pd$xA <- xA0*exp(-a*pd$t)
pd$xB <- -a * xA0 / (a - b) * exp(-a*pd$t) + a * xA0 / (a - b) * exp(-b*pd$t) + xB0*exp(-b*pd$t)
pd$xC <- w / (a - cc) * exp(-a*pd$t) +
  -w / (a - cc) * exp(-cc*pd$t) +
  -w / (b - cc) * exp(-b*pd$t) +
  w / (b - cc) * exp(-cc*pd$t) +
  -g / (b - cc) * exp(-b*pd$t) +
  g / (b - cc) * exp(-cc*pd$t) +
  xC0 * exp(-cc*pd$t)

plot_data <- melt(pd, id = "t")
g <- ggplot(plot_data, aes(x = t, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Salt (kg)", col = "Bucket")
plot(g)
```

## SIR Model

- **S**: Susceptible
- **I**: Infected + Infectious
- **R**: Removed (recovered, dead, immune)
- $\beta$: Rate of spread (1/time), P(transmission) per unit time
- $1/\beta$: Mean time before infection
- $\gamma$: Rate of recovery (1/time)
- $1/\gamma$: Mean time to recovery

```{r SIR_diagram, echo=FALSE, fig.height=3, fig.width=5}
temp <- DiagrammeR::grViz(diagram = "
  digraph boxes_and {
    graph[overlap = true, fontsize = 10, rankdir = LR]

    node [shape = box] S; I; R

    S->I [label=<&beta;>]
    I->R [label=<&gamma;>]
  }")
temp
rm(temp) # helps to prevent "out of memory errors"

```

## Equations {.smaller}

The sum of the compartments is the total population (100%)

$$S(t) + I(t) + R(t) = 1$$

The rate of decrease in the Susceptible is proportional to the fraction of
the population who are infectious ($I/N$) and the number of susceptible.

$$\frac{dS}{dt} = -\beta I S$$

The rate of increase is the Infections is the rate of change in the Susceptible
minus the rate at which Infectious turn to recovered.

$$\frac{dI}{dt} = \beta I S - \gamma I$$

The rate of change of the Recovered is equal to the rate the Infectious turn to
Recovered.

$$\frac{dR}{dt} = \gamma I$$

## Infection Rate

$$\frac{dI}{dt} = \gamma I \left(\frac{\beta}{\gamma} S - 1 \right)$$
$$ = \gamma I \left(r_0 S - 1 \right)$$

Note that the Infected will decrease immediately if $r_0 < \frac{1}{S}$ which 
people treat as effectively $r_0 < 1$ since $S$ starts near 100%.

If $r_0 > 1$, then the Infected will grow before decreasing later.

## Code

An exact solution is possible for the SIR model, but not for more complicated
models.  Most researchers use an ODE solver like `deSolve::ODE` in `R`

```{r sir, echo=TRUE}
sir <- function(t, st, pa)
{
  dS <- -1 * pa$beta * st["I"] * st["S"]
  dI <- pa$beta * st["I"] * st["S"] - pa$gamma * st["I"]
  dR <- pa$gamma * st["I"]
  
  return(list(c(dS, dI, dR)))
}

params <- list(beta = 2.2/7.5, gamma = 1/7.5)
state <- c(S = 0.99, I = 0.01, R = 0)
times <- seq(0, 125, by = 1)

out <- deSolve::ode(y = state, times = times, func = sir, parms = params)
```

## Figure

```{r plotsir, echo=FALSE}
plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```

## Reproduction < 1

$$r_0 < 1$$

```{r repro1, echo=FALSE}
params <- list(beta = 1,
               gamma = 2)
state <- c(S = 0.90,
           I = 0.10,
           R = 0)

times <- seq(0, 10, by = 0.1)
out <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```

## Reproduction Number > 1

$$r_0 > 1$$

```{r repro2, echo=FALSE}
params <- list(beta = 5,
               gamma = 2)
state <- c(S = 0.90,
           I = 0.10,
           R = 0)

times <- seq(0, 10, by = 0.1)
out <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```


## Flattening the Curve

```{r flat, echo=FALSE}
# r0 = 3
params <- list(beta = 6,
               gamma = 2)
state <- c(S = 0.99,
           I = 0.01,
           R = 0)

times <- seq(0, 10, by = 0.01)
out1 <- deSolve::ode(y = state, times = times, func = sir, parms = params)

# r0 = 1.5
params <- list(beta = 3,
               gamma = 2)
state <- c(S = 0.99,
           I = 0.01,
           R = 0)

times <- seq(0, 10, by = 0.01)
out2 <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data1 <- as.data.frame(out)
plot_data2 <- as.data.frame(out2)
plot_data <- data.frame(time = c(plot_data1$time, plot_data2$time),
                        I = c(plot_data1$I, plot_data2$I),
                        label = c(rep("r_0 = 3.0", nrow(plot_data1)),
                                  rep("r_0 = 1.5", nrow(plot_data2))))
g <- ggplot(plot_data, aes(x = time, y = I, col = label, group = label)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Scenario")
plot(g)
```

## Graph Simulation {.smaller}

The `igraph` package also has an `sir` method:

```{r sir2, echo=TRUE}
# sample a graph with 100 vertices and 100 edges
set.seed(1976)
gr <- igraph::sample_gnm(100, 100, directed = FALSE, loops = FALSE)
# start the epidemic at each node and simulate from there
sm <- igraph::sir(gr, beta = 5, gamma = 2)
plot(gr)
```

## Graph simulation figure

```{r sir2_figure, echo=FALSE}
plot(sm)
```

## Agent Based Model {.smaller}

```{r abm, echo=TRUE}
N <- 1000
state <- factor(c(rep("S", N - 1), rep("I", 1)), levels = c("S", "I", "R"))
number_contacts_per_time <- 2
prob_infect_given_contact <- 0.50
prob_recover_per_time <- 1/5
tabs <- vector("list", 60)
set.seed(1976)
for (i in seq_along(tabs))
{
  tabs[[i]] <- table(state)
  prob_contact_with_infected <- pmin(number_contacts_per_time * tabs[[i]]["I"] / N , 
                                     (N - 1)/N)
  contact_w_i <- rbinom(N, size = 1, prob = prob_contact_with_infected)
  infect <- ifelse(state == "S" & contact_w_i == 1, 
                   rbinom(N, 1, prob = prob_infect_given_contact), 
                   0)
  recover <- ifelse(state == "I", 
                    rbinom(N, 1, prob = prob_recover_per_time), 
                    0)
  state[which(infect == 1)] <- "I"
  state[which(recover == 1)] <- "R"
}
```

## ABM Figure

```{r plotabm, echo=FALSE}
plot_data <- as.data.frame(do.call("rbind", tabs))
plot_data$time <- seq_along(tabs)
plot_data <- melt(plot_data, id = "time")

g <- ggplot(plot_data, aes(x = time, y = value, col = variable, group = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population", col = "State")
plot(g)
```


## SEIR Model

- **S**: Susceptible
- **E**: Exposed and Infected or Incubating
- **I**: Infectious
- **R**: Removed (recovered, dead, immune)
- $\beta$: Rate of spread (1/time), P(transmission) per unit time
- $1/\beta$: Mean time before infection
- $\sigma$: Incubation rate (1/time)
- $1/\sigma$: Mean incubation time
- $\gamma$: Rate of recovery (1/time)
- $1/\gamma$: Mean time to recovery

## SEIR Model (2)

```{r SEIR_diagram, echo=FALSE}
temp <- DiagrammeR::grViz(diagram = "
  digraph boxes_and {
    graph[overlap = true, fontsize = 10, rankdir = LR]

    node [shape = box] S; E; I; R

    S->E [label=<&beta;>]
    E->I [label=<&sigma;>]
    I->R [label=<&gamma;>]
  }")
temp
rm(temp) # helps to prevent "out of memory errors"
```

## Equations

$$\frac{dS}{dt} = -\beta I S$$

$$\frac{dE}{dt} = \beta I S - \sigma E$$

$$\frac{dI}{dt} = \sigma E - \gamma I$$

$$\frac{dR}{dt} = \gamma I$$

## Code

```{r seir, echo=TRUE}
seir <- function(t, st, pa)
{
  dS <- -1 * pa$beta * st["I"] * st["S"]
  dE <- pa$beta * st["I"] * st["S"] - pa$sigma * st["E"]
  dI <- pa$sigma * st["E"] - pa$gamma * st["I"]
  dR <- pa$gamma * st["I"]
  return(list(c(dS, dE, dI, dR)))
}

# estimates from https://www.who.int/bulletin/online_first/20-255695.pdf
params <- list(beta = 2.2/7.5, gamma = 1 / 7.5, sigma = 1 / 2.3)
state <- c(S = 0.98, E = 0.01, I = 0.00, R = 0.00)
times <- seq(0, 175, by = 1)

out <- deSolve::ode(y = state, times = times, func = seir, parms = params)
```

## Figure

```{r plotseir, echo=FALSE}
plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time (days)", y = "Population Fraction", col = "Compartment")
plot(g)
```
