---
title: "SEIR Modeling"
author: "Rob Carnell"
date: "April 6, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(deSolve)
require(ggplot2)
require(reshape2)
```

## Basics

SIR Model

- S: Susceptible
- I: Infected + Infectious
- R: Removed (dead or immune)

SEIR Model

- S: Susceptible
- E: Exposed and Infected
- I: Infectious
- R: Removed (dead or immune)

## SIR Model

The sum of the compartments is the total population (100%)

$$S(t) + I(t) + R(t) = 1$$

The rate of decrease in the Susceptible is proportional to the fraction of
the population who are infectious ($I/N$) and the number of susceptible.

$$\frac{dS}{dt} = -\beta I S$$

The rate of increase is the Infections is the rate of change in the Susceptible
minus the rate at which Infectious turn to recovered.

$$\frac{dI}{dt} = \beta I S - \gamma I$$

$$ = \gamma I \left(\frac{\beta}{\gamma} S - 1 \right)$$
$$ = \gamma I \left(r_0 S - 1 \right)$$

The rate of change of the Recovered is equal to the rate the Infectios turn to
recovered.

$$\frac{dR}{dt} = \gamma I$$

```{r sir, echo=FALSE}
sir <- function(t, st, pa)
{
  dS <- -1 * pa$beta * st["I"] * st["S"]
  dI <- pa$beta * st["I"] * st["S"] - pa$gamma * st["I"]
  dR <- pa$gamma * st["I"]
  return(list(c(dS, dI, dR)))
}

params <- list(beta = 5,
               gamma = 2)
state <- c(S = 0.99,
           I = 0.01,
           R = 0)

times <- seq(0, 10, by = 0.1)
out <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```

## Reproduction number

$$r_0 < 1$$

```{r repro1, echo=FALSE}
params <- list(beta = 1,
               gamma = 2)
state <- c(S = 0.90,
           I = 0.10,
           R = 0)

times <- seq(0, 10, by = 0.1)
out <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```

$$r_0 > 1$$

```{r repro2, echo=FALSE}
params <- list(beta = 5,
               gamma = 2)
state <- c(S = 0.90,
           I = 0.10,
           R = 0)

times <- seq(0, 10, by = 0.1)
out <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data <- as.data.frame(out)
plot_data <- melt(plot_data, id = "time")
g <- ggplot(plot_data, aes(x = time, y = value, group = variable, col = variable)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Compartment")
plot(g)
```


## Flattening the Curve

```{r flat, echo=FALSE}
params <- list(beta = 6,
               gamma = 2)
state <- c(S = 0.99,
           I = 0.01,
           R = 0)

times <- seq(0, 10, by = 0.1)
out1 <- deSolve::ode(y = state, times = times, func = sir, parms = params)

params <- list(beta = 3,
               gamma = 2)

times <- seq(0, 10, by = 0.1)
out2 <- deSolve::ode(y = state, times = times, func = sir, parms = params)

plot_data1 <- as.data.frame(out)
plot_data2 <- as.data.frame(out2)
plot_data <- data.frame(time = c(plot_data1$time, plot_data2$time),
                        I = c(plot_data1$I, plot_data2$I),
                        label = c(rep("r_0 = 3.0", nrow(plot_data1)),
                                  rep("r_0 = 1.5", nrow(plot_data2))))
g <- ggplot(plot_data, aes(x = time, y = I, col = label, group = label)) +
  geom_line() +
  labs(x = "Time", y = "Population Fraction", col = "Scenario")
plot(g)
```

